Type theory in type theory with single substitutions

### 2024-09-07 types_2024_postproc/p.tex
stuff

\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\ext}{\mathop{\triangleright}}

@@

What is type theory? Here we refer to type theory as a particular
formal system based on Martin-Löf's original definition
\cite{martinlof73predicative}, and not to the field studying type
systems (e.g.\ \cite{DBLP:books/daglib/0005958}). When
answering this question, we aim for simplicity and ease of
understanding. We only require a working knowledge of an
implementation of type theory such as Agda. Our metatheory in which
this paper is written is thus type theory, and everything is
formalised in Agda.

The traditional way of defining a programming language formally
\cite{DBLP:books/cu/Ha2016,Pierce:SF2,plfa22.08} is comprised of three
steps: (i) definition of the presyntax (abstract syntax trees, ASTs)
as a formal grammar; (ii) specification of
well-formedness/well-typedness relations saying which ASTs are
well-behaved; (iii) specification of rewriting/conversion relations
explaining how programs are executed. Instead of following this route,
we use the intrinsic approach
\cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16}: (i)
we define ASTs as an inductive type instead of production rules of a
grammar; (ii) terms are indexed by (object theoretic) types, thus only
allowing ASTs to be built in well-typed ways; (iii) terms are
quotiented by conversion rules, making convertible terms equal. The
advantage of the intrinsic approach that it provides shorter, easier
to understand, less technical definitions, and there is no separation
between syntax and semantics. Type theory is a generalised algebraic
theory (GAT \cite{DBLP:journals/apal/Cartmell86}), the notion of model
is direct, and the syntax is simply the initial model. GATs are
generalisations of single-sorted algebraic theories (such as monoids,
groups) to multiple sorts where later sorts can be indexed over
previous sorts. A typical example is categories with a sorts of
objects and a sort of morphisms double-indexed over objects. The
initial model of a GAT is a quotient inductive-inductive type (QIIT
\cite{DBLP:journals/pacmpl/KaposiKA19}). QIITs are a general family of
inductive types where multiple types are defined mutually, later types
can be indexed over previous ones, and equality constructors are also
allowed, generalising quotients. The elimination principle of the QIIT
corresponds to a structural induction principle for the QIIT and is
equivalent to the statement that the syntax is the initial model
\cite{DBLP:journals/pacmpl/KaposiKA19}, that is, there is a unique
morphism from the syntax to any other model.

There are several ways to define type theory as a GAT. For example,
Thomas Ehrhard's calculus \cite{ehrhard,coquandEhrhard}, Dybjer's
categories with families (CwFs
\cite{DBLP:conf/types/Dybjer95,Castellan2021}), contextual categories
or C-systems
\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18},
natural models \cite{DBLP:journals/mscs/Awodey18}, B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}. These are providing the
basic substitution calculus part of the theory, and individual type
formers (such as function space or product types) have to be specified
separately. With the exception of B-systems, the listed definitions
are all based on parallel substitution calculi, where contexts and
substitutions form a category. Like B-systems, in this paper we define
type theory using \emph{single substitutions}. Single substitutions
don't compose, so they do not form a category. Our definition is
simpler than B-systems, but it is closely related to it. It is a
minimalistic presentation where every operation and equation is
well-motivated (see Section \ref{sec:ssc}). In addition to its
pedagogical value due to its simplicity, our definition might
contribute in the future to a possible coherent syntax for type
theory, see Section \ref{sec:conclusion}.

Our theory is in fact a bit too minimalistic to be workable: for
example, consider a term $b$ which depends on a context $\Gamma$ and
an extra variable of type $A$, that is, $b : \Tm\,(\Gamma\ext
A)\,B$. We can weaken $b$ so that we add an extra $A$ variable
\emph{before} the last one obtaining $b[\p^+] : Tm\,(\Gamma\ext A\ext
A[\p])\,(B[\p])$ where $\p$ is the weakening substitution. Then, we
instantiate the last $A[\p]$ variable in the context by the previous
$A$ obtaining $b[\p^+][\langle\q\rangle]$ where $\q$ is De Bruijn
index $0$. We cannot derive $b[\p^+][\langle\q\rangle] = b$ in an
arbitrary model, but this equation is admissible in the \emph{syntax}.
% p⁺  : Sub (Γ▹A▹A[p]) (Γ▹A)
% ⟨q⟩ : Sub (Γ▹A) (Γ▹A▹A[p])
This is analogous to e.g.\ parametricity results
\cite{DBLP:journals/jfp/BernardyJP12}, which do not hold in an
arbitrary model, but they hold for syntactic terms. In fact, all the
operations and equations of the parallel substitution calculus (CwF)
are admissible in our syntax, moreover the set of single substitution
based syntactic terms are isomorphic to the set of CwF-based
terms. Every CwF-based model is also a model of our single
substitution calculus, but not the other way: there are more
SSC-models than CwF-models. The situation is analogous to that of
extensional combinatory calculus and lambda calculus
\cite{DBLP:conf/fscd/AltenkirchKSV23}: the syntactic terms form
isomorphic sets, but there are more combinatory models than lambda
calculus models.

TODO: rewrite from here. Add: discussion of type formers, SOGATs and say that we only do this
for an example t.t. and not for all of them. We conjecture that for
any SOGAT, the SSC and CwF-based syntaxes are equivalent.

The summary of the above discussion: we define a new minimal syntax
for type theory which is equivalent to the original syntax, but

@@

Our main contribution is the novel definition of a minimalistic single
substitution calculus, the $\alpha$-normalisation algorithm which can
be useful for other explicit substitution calculi, and finally our
proof of equivalence with the CwF-style syntax.

@@

\section{The single substitution syntax is a category with families}

@@

@@

SSC was used in the 90s for simple type theories, ask Hugo for references.

@@

Maybe we can do without the extra type equations, by using assumptions.

Luksa, Thorsten thesis.

>>>>

\newcommand{\ra}{\rightarrow}
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\ext}{\mathop{\triangleright}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\lam}{\mathsf{lam}}
\newcommand{\app}{\mathsf{app}}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\cd}{\mathsf{c}}

@@

What is type theory? Here we refer to type theory as a particular
formal system based on Martin-Löf's original definition
\cite{martinlof73predicative}, and not to the field studying type
systems (e.g.\ \cite{DBLP:books/daglib/0005958}). When answering this
question, we aim for simplicity and ease of understanding.

The traditional way of defining a programming language formally
\cite{DBLP:books/cu/Ha2016,Pierce:SF2,plfa22.08} is comprised of three
steps: (i) definition of the presyntax (abstract syntax trees, ASTs)
as a formal grammar; (ii) specification of
well-formedness/well-typedness relations saying which ASTs are
well-behaved; (iii) specification of rewriting/conversion relations
explaining how programs are executed. Instead of following this route,
we use the intrinsic approach
\cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16}: (i)
we define ASTs as an inductive type instead of production rules of a
grammar; (ii) terms are indexed by (object theoretic) types, thus only
allowing ASTs to be built in well-typed ways; (iii) terms are
quotiented by conversion rules, making convertible terms equal. The
advantage of the intrinsic approach is that it provides shorter,
easier to understand, less technical definitions, and there is no
mismatch between syntax and semantics. Type theory is a generalised
algebraic theory (GAT \cite{DBLP:journals/apal/Cartmell86}), the
notion of model is direct, and the syntax is simply the initial
model. GATs are generalisations of single-sorted algebraic theories
(such as monoids, groups) to multiple sorts where later sorts can be
indexed over previous sorts. An example is categories with a sort of
objects and a sort of morphisms double-indexed over objects. In a
type-theoretic metatheory, the initial model of a GAT is a quotient
inductive-inductive type (QIIT
\cite{DBLP:journals/pacmpl/KaposiKA19}). QIITs are a general family of
inductive types where multiple types are defined mutually, later types
can be indexed over previous ones, and equality constructors are also
allowed, generalising quotients. The elimination principle of the QIIT
corresponds to a structural induction principle and is
equivalent to the statement that the syntax is the initial model
\cite{DBLP:journals/pacmpl/KaposiKA19}, that is, there is a unique
morphism from the syntax to any other model.

There are several ways to define type theory as a GAT. For example,
Thomas Ehrhard's calculus \cite{ehrhard,coquandEhrhard}, Dybjer's
categories with families (CwFs
\cite{DBLP:conf/types/Dybjer95,Castellan2021}), contextual categories
or C-systems
\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18},
natural models \cite{DBLP:journals/mscs/Awodey18}, B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}. These are providing the
basic substitution calculus part of the theory, and individual type
formers (such as function space or product types) have to be specified
separately. With the exception of B-systems, the listed definitions
all define a parallel substitution calculus, where contexts and
substitutions form a category. Like B-systems, in this paper we define
type theory as a single substitution calculus (SSC). Single
substitutions don't compose, so they do not form a category. Our
definition is simpler than B-systems, but it is closely related to
it. It is a minimalistic presentation where every operation and
equation is well-motivated (see Section \ref{sec:ssc}). In addition to
its pedagogical value due to its simplicity, our definition might
contribute in the future to a possible coherent syntax for type
theory, see Section \ref{sec:conclusion}.

Our theory is in some sense too minimalistic: for example, consider a
term $b$ which depends on a context $\Gamma$ and an extra variable of
type $A$, that is, $b : \Tm\,(\Gamma\ext A)\,B$. We can weaken $b$ so
that we add an extra $A$ variable \emph{before} the last one obtaining
$b[\p^+] : \Tm\,(\Gamma\ext A\ext A[\p])\,(B[\p])$ where $\p$ is
weakening at the end of the context, and $\p^+$ is weakening just
before the last variable in the context. Then, we instantiate the last
$A[\p]$ variable in the context by the previous $A$ obtaining
$b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$ where $\q$ is De
Bruijn index $0$. We cannot derive $b[\p^+][\langle\q\rangle] = b$ in
an arbitrary model, but this equation holds in the syntax.
% p⁺  : Sub (Γ▹A▹A[p]) (Γ▹A)
% ⟨q⟩ : Sub (Γ▹A) (Γ▹A▹A[p])
This is analogous to e.g.\ parametricity results
\cite{DBLP:journals/jfp/BernardyJP12}, which do not hold in an
arbitrary model, but they hold for syntactic terms. In fact, all the
operations and equations of the parallel substitution calculus (CwF)
are admissible in our syntax, moreover the set of SSC-based syntactic
terms are isomorphic to the set of CwF-based terms. In contrast with
B-systems which are equivalent to CwFs, our SSC-models do not form
CwF-based models. Every CwF-based model is a model of our SSC, but not
the other way: there are more SSC-models than CwF-models. The
situation is analogous to the relationship of extensional combinatory
calculus and lambda calculus \cite{DBLP:conf/fscd/AltenkirchKSV23}:
the syntactic terms form isomorphic sets, but there are more
combinatory models than lambda calculus models.

In this paper, we define the SSC and show how it can be extended with
$\Pi$ types and a hierarchy of Coquand-universes
\cite{coquandUniverse}. The rules for type formers are the same as for
CwFs, the difference is only in the substitution calculus part. More
generally, every second-order generalised algebraic theory (SOGAT) has
a parallel and a single substition-based first order presentation
\cite{DBLP:conf/fscd/KaposiX24}. Here we prove that the syntaxes of
the parallel and the single substitution calculi are isomorphic for
the following SOGAT (we use implicit arguments and $\cong$ denotes
isomorphism):
\begin{alignat*}{10}
  & \Ty && : \N\ra\Set \\
  & \Tm && : \Ty\,i\ra\Set \\
  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
  & \U && : (i:\N)\ra\Ty\,(1+i) \\
  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd
\end{alignat*}
We conjecture that the methods of in this paper can be used to show
that for any SOGAT, the two first-order presentations give isomorphic
syntaxes.

TODO: say something about alpha-normalisation.

To summarise, our contributions are the following:
\begin{itemize}
\item A new generalised algebraic presentation of type theory in the
  form of a minimalistic single substitution calculus.
\item The $\alpha$-normalisation technique which is useful for working
  with explicit substitution calculi.
\item The proof of equivalence with the parallel, CwF-based syntax.
\end{itemize}

In Section \ref{sec:ssc}, we introduce

@@

We only require a working knowledge of an implementation of type
theory such as Agda. Our metatheory in which this paper is written is
thus type theory, and everything is formalised in Agda.

\subsection{Related work}

Algebraic formalisations of type theory: Brunerie
\cite{initiality-agda} \cite{DBLP:conf/popl/AltenkirchK16},
normalisation, shallow embedding canonicity, gluing. Low level
formalisations: Abel, Loic, metacoq. They can be seen as unfolding the
more abstract thing in a (weak) setoid model.

Other SSC: 90s French, Thorsten thesis, B-systems, Luksa paper.

@@

Show examples of using the calculus, and where we need to derive more
stuff.

\section{Deriving parallel substituions}

@@

\section{Isomorphism}

@@

@@

Separate sort for variables, without the extra type equations, by
using assumptions, and proving everything by induction.
